Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-11-28 analysis/paper/paper.Rmd
Additional analyses after PCI review

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Department of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: true
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
  \usepackage[left]{lineno}
  \linenumbers
# always_allow_html: true
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  By developing a new methodology for handling and assessing a large number of shoreline dated sites, this paper compares the summed probability distribution of radiocarbon dates and shoreline dates along the Skagerrak coast of south-eastern Norway. Both measures have previously been compared to elucidate demographic developments in Fennoscandia, but these have not been based on probabilistic methods for shoreline dating. The findings indicate a largely diverging development of the two data sets through the Mesolithic. The number of shoreline dated sites undergoes some process of overall decrease through the period, while the radiocarbon data is characterised by a lacking signal in the earliest parts of the period and then undergoes a logistic growth that quickly plateaus and remains stable for the remainder of the period. Although the precise nature of this discrepancy will require further substantiation, we tentatively suggest that while not devoid of a demographic signal, the number of shoreline dated sites is heavily influenced by mobility patterns. Conversely, we also suggest that the lacking signal in the radiocarbon data for the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data more directly reflects population dynamics.
keywords: | 
  Mesolithic; Shoreline dates; Radiocarbon dates; Fennoscandia
---
Keywords: `r rmarkdown::metadata$keywords`

@@

The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE have a likely elevation above sea-level that can be reasonably approximated by a gamma distribution with shape =  0.286 and scale = 20.833. This is incorporated in the procedure for shoreline dating used for this study (Figure \@ref(fig:shoredate)D). When performing this dating procedure, the gamma distribution is sequentially stepped through and transferred to the calendar scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve. This procedure thus gives the shoreline date of a site when accounting for the likely elevation of the site above sea-level when it was in use [see @roalkvam2023 for details]. Given that the shoreline displacement curves have no inversions and should therefore before commutative [cf. @weninger2015, 545], each shoreline date is normalised to sum to unity before being summed. To reduce the computational cost of the simulation procedures to follow below, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.

@@

All the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. Its negative direction indicates that some overall process of reduction in the frequency of shoreline-dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Figure \@ref(fig:spdmc)A). Some significant deviations occur with the model over-predicting the frequency of sites in the period from c. 9200--9000 BCE, followed by a significant positive peak just after 9000 BCE, lasting until around 8500 BCE. The process of decline is then interrupted by positive deviations at 7900--7400, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation between c. 6800 and 5800 BCE. Finally, a positive deviation occurs again around 5300--5000 BCE. While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded in the discussion to follow.

@@

Although there are generally few ^14^C-dates from the earliest part of the Mesolithic in Norway [e.g. @åstveit2018; @kleppe2018; @breivik2018a; @solheim2018], it is established that Stone Age sites in southern Norway date to the period from before the earliest ^14^C-dates in the study area for this paper, and at least from as far back as around 9300 BCE [e.g. @damlien2018b; @glorstad2016]. While the SSPD achieved here starts from c. 9500 BCE, the RSPD shows no evidence of human activity until c. 8500 BCE. What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years. These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @åstveit2014; @bjerck2017; @damm2022]. A possible lack of reliable access to firewood has also been seen as possible contributor to this pattern, although evidence for the early presence of forested landscapes, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; cf. @damm2022; @sørensen2014]. While taphonomic and ecological factors are likely part of the picture, we view it as probable that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind this pattern.

@@

A common interpretation of the Early Mesolithic in Norway is that the population was small but highly mobile, reflected in large numbers of short-lived sites along the coast [e.g. @bjerck2008; @bjerck2017; @fuglestvedt2012]. This is commonly seen in relation to homogeneous lithic technology over large swathes of the Scandinavian Peninsula in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline [e.g. @bang-andersen2012; @bjerck2009; @fuglestvedt2012]. High mobility has been seen as necessary both to maintain demographic viability among colonising groups, and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017; also discussed by @berg-hansen2018]. The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period. The lacking signal in the RSPD for the first settlement period could in this view reflect mobility patterns characterised by short occupational duration and corresponding pyrotechnology based on small, short-lived fires [cf. @damm2022] and low population numbers.

@@

A recent study by @manninen2023, which considers a larger inland area of northern Scandinavia than the above-referenced studies, found a corresponding increase in inland activity starting from the earliest parts of the Mesolithic that lasted until c. 6000 BCE. However, after this the population dynamics indicated by the radiocarbon record is found to have been characterised by a cessation of growth, rather than the clear decrease suggested by the regional studies in southern Norway. Consequently, properly relating the developments in the coastal region to inland developments and their relevance to each other will benefit from an integrated approach that also explicitly considers the spatial dimension without invoking predefined analytical regions [cf. @crema2017].

Following from the divergence between the proxies, assessing the suggested and possibly varying influence of demographic developments and mobility patterns on the measures will remain speculative without drawing on additional variables that can control for such confounding effects. Following from the stationarity of the logistic model of the RSPD after c. 7500 BCE, it is possible that fluctuations in the SSPD reflect variation in mobility patterns that are unrelated to demographic developments. There does appear to be a negative deviation from the logistic null model in the RSPD corresponding to the one in the SSPD just before 6000 BCE. However, the logistic model could not be rejected for the RSPD, and so it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, at least given the present data. While it is possible that the limited sample of ^14^C-dates has led to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018]. A related point that is also worth making here is that the overall demographic trends that the modelling efforts can hope to identify are likely to have been subject to smaller scale demographic fluctuations that simply cannot be ascertained given the available data and its resolution [cf. @tallavaara2021], and which impacts the comparison of the two proxies following both from the different resolutions of the dating methods and the sample sizes underlying the SPDs.

@@

Some further lines of inquiry could be to draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @barton2014; @clark2017; @preston2018; @roalkvam2022], which is potentially also influenced by variable rates of relative sea-level change. In terms of mobility patterns, focus was directed here towards residential movement and occupational duration; but other dimensions relating for example to seasonal and territorial mobility, as reflected in the discussion of inland activity, could also prove to be important. In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g. @hoebe2023; @manninen2018; @jorgensen2020; @ordonez2022; @lundstrom2023]. To account for the confounding effects these might have, a more principled and comprehensive evaluation of how investigatory practices might impact the SPDs would also be beneficial. While certainly also pertinent for the RSPD, this has generally been considered more extensively for radiocarbon data than for shoreline dated sites. Furthermore, aligning the complex web of causes and effects that these different lines of evidence potentially represent will also benefit from a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf. @timpson2021]. Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data employed here, but could also provide important contextualisation for the observed development [see e.g. @lundstrom2021; @manninen2023; @solheim2018] and potentially reduce the impact of idiosyncratic investigations that might impact the present SPDs inordinately [see e.g. @crema2022; @hoebe2023; @rick1987].

>>>>

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Department of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: true
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
  \usepackage[left]{lineno}
  \linenumbers
# always_allow_html: true
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  By developing a new methodology for handling and assessing a large number of shoreline dated sites, this paper compares the summed probability distribution of radiocarbon dates and shoreline dates along the Skagerrak coast of south-eastern Norway. Both measures have previously been compared to elucidate demographic developments in Fennoscandia, but these have not been based on probabilistic methods for shoreline dating. The findings indicate a largely diverging development of the two data sets through the Mesolithic. The number of shoreline dated sites undergoes some process of overall decrease through the period, while the radiocarbon data is characterised by a lacking signal in the earliest parts of the period and then undergoes a logistic growth that quickly plateaus and remains stable for the remainder of the period. Although the precise nature of this discrepancy will require further substantiation, we tentatively suggest that while not devoid of a demographic signal, the number of shoreline dated sites could be more heavily influenced by mobility patterns. Conversely, we also suggest that the lacking signal in the radiocarbon data for the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data more directly reflects population dynamics.
keywords: | 
  Mesolithic; Shoreline dates; Radiocarbon dates; Fennoscandia
---
Keywords: `r rmarkdown::metadata$keywords`

@@

The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE have a likely elevation above sea-level that can be reasonably approximated by a gamma distribution with shape =  0.286 and scale = 20.833. This is incorporated in the procedure for shoreline dating used for this study (Figure \@ref(fig:shoredate)D). When performing this dating procedure, the gamma distribution is sequentially stepped through and transferred to the calendar scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve. This procedure thus gives the shoreline date of a site when accounting for the likely elevation of the site above sea-level when it was in use [see @roalkvam2023 for details]. Given that the shoreline displacement curves have no inversions and should therefore be commutative [see @weninger2015, 545], each shoreline date is normalised to sum to unity before being summed. To reduce the computational cost of the simulation procedures to follow below, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.

@@

All the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. Its negative direction indicates that some overall process of reduction in the frequency of shoreline-dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Figure \@ref(fig:spdmc)A). Some significant deviations occur with the model over-predicting the frequency of sites in the period from c. 9200--9000 BCE, followed by a significant positive peak just after 9000 BCE, lasting until around 8500 BCE. The process of decline is then interrupted by positive deviations at 7900--7400, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation between c. 6800 and 5800 BCE. Finally, a positive deviation occurs again around 5300--5000 BCE. While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [@roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded in the discussion to follow.

@@

Although there are generally few ^14^C-dates from the earliest part of the Mesolithic in Norway [e.g. @åstveit2018; @kleppe2018; @breivik2018a; @solheim2018], it is established that Stone Age sites in southern Norway date to the period from before the earliest ^14^C-dates in the study area for this paper, and at least from as far back as around 9300 BCE [e.g. @damlien2018b; @glorstad2016]. While the SSPD achieved here starts from c. 9500 BCE, the RSPD shows no evidence of human activity until c. 8500 BCE. What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years. These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @åstveit2014; @bjerck2017; @damm2022]. A possible lack of reliable access to firewood has also been seen as a possible contributor to this pattern, although evidence for the early presence of forested landscapes, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; see also @damm2022; @sørensen2014]. While taphonomic and ecological factors are likely part of the picture, we view it as probable that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind this pattern.

@@

A common interpretation of the Early Mesolithic in Norway is that the population was small but highly mobile, reflected in large numbers of short-lived sites along the coast [e.g. @bjerck2008; @bjerck2017; @fuglestvedt2012]. This is commonly seen in relation to homogeneous lithic technology over large swathes of the Scandinavian Peninsula in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline [e.g. @bang-andersen2012; @bjerck2009; @fuglestvedt2012]. High mobility has been seen as necessary both to maintain demographic viability among colonising groups, and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017; also discussed by @berg-hansen2018]. The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period. The lacking signal in the RSPD for the first settlement period could in this view reflect mobility patterns characterised by short occupational duration and corresponding pyrotechnology based on small, short-lived fires [@damm2022] and low population numbers.

@@

A recent study by @manninen2023, which considers a larger inland area of northern Scandinavia than the above-referenced studies, found a corresponding increase in inland activity starting from the earliest parts of the Mesolithic that lasted until c. 6000 BCE. However, after this the population dynamics indicated by the radiocarbon record is found to have been characterised by a cessation of growth, rather than the clear decrease suggested by the regional studies in southern Norway. Consequently, properly relating the developments in the coastal region to inland developments and their relevance to each other will benefit from an integrated approach that also explicitly considers the spatial dimension without invoking predefined analytical regions [see @crema2017].

Following from the divergence between the proxies, assessing the suggested and possibly varying influence of demographic developments and mobility patterns on the measures will remain speculative without drawing on additional variables that can control for such confounding effects. Following from the stationarity of the logistic model of the RSPD after c. 7500 BCE, it is possible that fluctuations in the SSPD reflect variation in mobility patterns that are unrelated to demographic developments. There does appear to be a negative deviation from the logistic null model in the RSPD corresponding to the one in the SSPD just before 6000 BCE. However, the logistic model could not be rejected for the RSPD, and so it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, at least given the present data. While it is possible that the limited sample of ^14^C-dates has led to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018]. A related point that is also worth making here is that the overall demographic trends that the modelling efforts can hope to identify are likely to have been subject to smaller scale demographic fluctuations that simply cannot be ascertained given the available data and its resolution [@tallavaara2021], and which impacts the comparison of the two proxies following both from the different resolutions of the dating methods and the sample sizes underlying the SPDs.

@@

Some further lines of inquiry could be to draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @barton2014; @clark2017; @preston2018; @roalkvam2022], which is potentially also influenced by variable rates of relative sea-level change. In terms of mobility patterns, focus was directed here towards residential movement and occupational duration; but other dimensions relating for example to seasonal and territorial mobility, as reflected in the discussion of inland activity, could also prove to be important. In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g. @hoebe2023; @manninen2018; @jorgensen2020; @ordonez2022; @lundstrom2023]. To account for the confounding effects these might have, a more principled and comprehensive evaluation of how investigatory practices might impact the SPDs would also be beneficial. While certainly also pertinent for the RSPD, this has generally been considered more extensively for radiocarbon data than for shoreline dated sites. Furthermore, aligning the complex web of causes and effects that these different lines of evidence potentially represent will also benefit from a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [see @timpson2021]. Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data employed here, but could also provide important contextualisation for the observed development [see e.g. @lundstrom2021; @manninen2023; @solheim2018] and potentially reduce the impact of idiosyncratic investigations that might impact the present SPDs inordinately [see e.g. @crema2022; @hoebe2023; @rick1987].
